Population structure and phenotypic variation of *Sclerotinia sclerotiorum* from dry bean in the United States

### 2017-07-12 doc/manuscript/manuscript.Rmd
Comment out current discussion

Our study was conducted to characterize *S. sclerotiorum* isolates from different geographic areas to determine important phenotypic variables influencing genotypes; to ascertain reproductive behavior of the pathogen; to test whether different sampling years or locations have any effect on genotype/sexuality; and to identify abundant and unique genotypes across isolates and geographies.
The dbRDA revealed region, year, and host as the most influential phenotypic variables on genotypes.
However, further molecular analysis of pathogen isolates from selected hosts did not show any significant genetic difference among populations.
According to dbRDA, aggressiveness and location of isolates had little or no effect.
Previous studies on *S. sclerotiorum* from alfalfa soil [@attanayake2012genetic] and potato [@atallah2004high], and canola [@attanayake2013sclerotinia] plants reported little or no correlation of genotypes on pa25thogen aggressiveness.
This is important information for resistant breeding since genotypes of *S. sclerotiorum* isolates seem to depend to a certain degree on different levels of region and also on environmental conditions of the sampling year.
For instance our analysis indicated that MX, 2009, and 2005 were highly influential predictors of *S. sclerotiorum* genotypes (Fig. 7).
Conversely, France and Australia, which are geographically distant to North America, were not within the top 10 predictors of genetic composition of isolates (Fig. 7).
According to dbRDA, Mexican region imposed the highest impact on microsatellite loci of *S. sclerotiorum* isolates.
We used Bruvo's distance for calculating genetic distances of different isolates since it incorporates stepwise mutation model and applicable to microsatellite analysis.

@@

Our study did not find any multilocus genotype/s common across all or majority of sampling geographies. <!-- ZNK: I think this statement is a bit misleading. Yes, we didn't find any individual MLGs, but 46% of these are present across at least two populations.  -->
The most widespread and common genotype (MLG 25, Supplemental Table 1) was found only in five states of the USA with majority of isolates confined to ND.
However, we did find some MLGs of France and Australian isolates to be present in the USA samples.
Some studies on *S. sclerotiorum* did not detect any shared MLGs between geographically distant populations.
@attanayake2013sclerotinia were unable to detect any shared microsatellite MLGs in *S. sclerotiorum* populations from infected canola fields in China and USA.
Isolate collections were limited to a single county in China whereas USA isolates were from ND state only.
This may have limited the chance of discovering genotypes common to both regions.
@atallah2004high examined the population structure of *S. sclerotiorum* in different potato fields of Columbia basin and reported only 15 isolates shared the most common microsatellite haplotype in a sample of 167.
Similarly, several other studies did not detect common fingerprints of *S. sclerotiorum* isolates in geographically distant populations: Australia and North America [@ekins1999genetic];  California, Washington, and Ontario [@malvarez2007new]; Canada, North Carolina, and Louisiana [@cubeta1997clonality].

@@

Interestingly, all Mexican genotypes were unique to that region.
On the other hand CA genotypes were present in the West Coast states of OR and WA which may be due to movement of sclerotia with planting material. <!-- ZNK: Do we have some sort of evidence for this, or is this pure speculation? -->
<!-- 
I'm not sure what the point of this paragraph is.
I believe I can put the discussion of the graph results here. 
-->
Surprisingly, France and Australian genotypes were also isolated in the USA.
MLG 4 (see supplemental Table 1) was recorded from France, Australia and MN while France and USA isolates had four common MLGs.
Similarly, five MLGs were recorded common to both Australia and USA isolates.
However, there were six unique genotypes in clone-corrected France population (N= 6/14) and three genotypes in CA population (N = 3/15).
MLGs from Australian region, our samples were limited to 6, which is not a sufficient number to make inferences.
Our results indicated that different continents were not a defining factor for microsatellite genotypes of *S. sclerotiorum* isolated from dry bean.
This may be due to involuntary movement of sclerotia by human activities and probably because of recent speciation of *S. sclerotiorum*. 
A phylogenetic study on Sclerotinaceae showed that most of the sclerotial lineages including different species of Sclerotinia had evolved relatively recently from a common ancestor [@carbone1993ribosomal].
Few multilocus genotypes were unique to each sampling state of the USA but most of them were common among several states (Supplemental Table 1).
For instance, Michigan state from which we had the largest collection of isolates (N = 62) consisted of 43 MLGs.
Of these, 32 MLGs were present in at least one of the 5 states of NE, ND, WA, CO, MN; 4 MLGs were present in France isolates; and 11 were unique to MI sites (see Supplemental Table 1 for unique MLGs in different regions).
When selecting *S. sclerotiorum* isolates for screening dry bean germplasm for disease resistance, plant breeders may find above information useful.
A study on genetic diversity of *S. sclerotiorum* from multiple host crops in the North central region of the USA revealed there was little evidence for genetic isolation of this pathogen by host crop or by distance [@aldrich-wolfe2015genetic].
Though our isolates didn't come from different crops, we had isolates from many resistant and susceptible dry bean cultivars, which also didn't show genetic differentiation according to AMOVA.
It seems Mexican pathogen isolates are different from the rest and warrants further investigation using a larger population.
Some France genotypes were common in USA isolates but at the same time 43% (6/14) of the France genotypes were unique to that region indicating white mold resistant cultivars produced in the USA might not work in that region. <!-- ZNK: I'm not entirely sold on this conclusion. -->
Geographic distance between different states of the USA seem not a barrier for gene transfer of *S. sclerotiorum* isolates though unique genotypes did prevail at a low frequency in each state. <!-- ZNK: I don't know that this adds anything. -->
The genetic diversity of regions with >10 isolates had similar values (0.43 - 0.59) except for Mexican isolates.
*h* of MX was 0.26 and indicated less exposure of sampling sites to outside inocula. <!-- ZNK: Does this mean that there's heavy inbreeding here? -->
Test for linkage disequilibrium showed evidence of clonal reproduction for MX isolates (Ia = 2.61, P.Ia = 0.001, $\bar{r}_d$ = 0.37, P.$\bar{r}_d$  = 0.001).
This may have been another reason for low gene diversity.

DAPC analysis by region followed by AMOVA corroborated the uniqueness of Mexican genotypes by grouping them away from the rest of the isolates (Fig. 8).
Other populations based on geography were unable to group in separate clusters except for the Californian population.
We tested whether genotypes of *S. sclerotiorum* populations were influenced by different dry bean host cultivars and year of sampling by DAPC and AMOVA.
There was no host specificity but year of sampling did show a significant impact on genotypes.
Other studies also reported host identity is no barrier for gene flow of this pathogen [@aldrich-wolfe2015genetic; @clarkson2013population].
@atallah2004high revealed statistically significant impact of the collection date of *S. sclerotiorum* isolates from four potato fields in the Columbia Basin on molecular variance.
But there was no impact of the year or location of collection.
We also did not detect location as a significant predictor of molecular variance but year had a significant impact.
Our samples were collected on a single day in each year from each field and therefore did not warrant for an analysis of date of sample collection. <!-- BSA: Is this true, coauthors? -->

@@

*Sclerotinia sclerotiorum* isolates from dry bean represented high microsatellite diversity and lack of association MCGs and microsatellite MLGs.
Analysis of molecular variance of populations based on region and sampling year showed low level of differentiation with majority of variance detected within populations.
These features are indicative of sexual outcrossing and perhaps human mediated migration of *S. sclerotiorum* in dry bean.
However, we also detected clones or identical microsatellite MLGs within and among populations indicating asexual reproduction.
Further, linkage disequilibrium and index of association calculations suggested *S. sclerotiorum* populations are largely clonal.
Although, there was low impact of region and year on population differentiation, we did not detect host cultivar as a barrier to gene flow of the pathogen.
Our study did not find a widespread genotype among sampled regions from dry beans but some Australian and France MLGs were found in the USA isolates.
The analysis of populations from same field over different years for linkage disequilibrium did not provide evidence for sexual outcrossing.
However, analysis of molecular variance did show populations were significantly different between some years (eg.
MI 2003 and 2005 populations; WA 2008 compared to both 2003 and 2005).
These results indicate some level of heterothallic recombination is taking place in these fields in addition to natural mutations.
An understanding of sexual behavior and ability of form new genotypes by *S. sclerotiorum* in dry bean cropping systems is useful for formulating disease management strategies.

>>>>

<!-- 
Our study was conducted to characterize *S. sclerotiorum* isolates from different geographic areas to determine important phenotypic variables influencing genotypes; to ascertain reproductive behavior of the pathogen; to test whether different sampling years or locations have any effect on genotype/sexuality; and to identify abundant and unique genotypes across isolates and geographies.
The dbRDA revealed region, year, and host as the most influential phenotypic variables on genotypes.
However, further molecular analysis of pathogen isolates from selected hosts did not show any significant genetic difference among populations.
According to dbRDA, aggressiveness and location of isolates had little or no effect.
Previous studies on *S. sclerotiorum* from alfalfa soil [@attanayake2012genetic] and potato [@atallah2004high], and canola [@attanayake2013sclerotinia] plants reported little or no correlation of genotypes on pa25thogen aggressiveness.
This is important information for resistant breeding since genotypes of *S. sclerotiorum* isolates seem to depend to a certain degree on different levels of region and also on environmental conditions of the sampling year.
For instance our analysis indicated that MX, 2009, and 2005 were highly influential predictors of *S. sclerotiorum* genotypes (Fig. 7).
Conversely, France and Australia, which are geographically distant to North America, were not within the top 10 predictors of genetic composition of isolates (Fig. 7).
According to dbRDA, Mexican region imposed the highest impact on microsatellite loci of *S. sclerotiorum* isolates.
We used Bruvo's distance for calculating genetic distances of different isolates since it incorporates stepwise mutation model and applicable to microsatellite analysis.

@@

Our study did not find any multilocus genotype/s common across all or majority of sampling geographies. <!-~ ZNK: I think this statement is a bit misleading. Yes, we didn't find any individual MLGs, but 46% of these are present across at least two populations.  ~->
The most widespread and common genotype (MLG 25, Supplemental Table 1) was found only in five states of the USA with majority of isolates confined to ND.
However, we did find some MLGs of France and Australian isolates to be present in the USA samples.
Some studies on *S. sclerotiorum* did not detect any shared MLGs between geographically distant populations.
@attanayake2013sclerotinia were unable to detect any shared microsatellite MLGs in *S. sclerotiorum* populations from infected canola fields in China and USA.
Isolate collections were limited to a single county in China whereas USA isolates were from ND state only.
This may have limited the chance of discovering genotypes common to both regions.
@atallah2004high examined the population structure of *S. sclerotiorum* in different potato fields of Columbia basin and reported only 15 isolates shared the most common microsatellite haplotype in a sample of 167.
Similarly, several other studies did not detect common fingerprints of *S. sclerotiorum* isolates in geographically distant populations: Australia and North America [@ekins1999genetic];  California, Washington, and Ontario [@malvarez2007new]; Canada, North Carolina, and Louisiana [@cubeta1997clonality].

@@

Interestingly, all Mexican genotypes were unique to that region.
On the other hand CA genotypes were present in the West Coast states of OR and WA which may be due to movement of sclerotia with planting material. <!-~ ZNK: Do we have some sort of evidence for this, or is this pure speculation? ~->
<!-~ 
I'm not sure what the point of this paragraph is.
I believe I can put the discussion of the graph results here. 
~->
Surprisingly, France and Australian genotypes were also isolated in the USA.
MLG 4 (see supplemental Table 1) was recorded from France, Australia and MN while France and USA isolates had four common MLGs.
Similarly, five MLGs were recorded common to both Australia and USA isolates.
However, there were six unique genotypes in clone-corrected France population (N= 6/14) and three genotypes in CA population (N = 3/15).
MLGs from Australian region, our samples were limited to 6, which is not a sufficient number to make inferences.
Our results indicated that different continents were not a defining factor for microsatellite genotypes of *S. sclerotiorum* isolated from dry bean.
This may be due to involuntary movement of sclerotia by human activities and probably because of recent speciation of *S. sclerotiorum*. 
A phylogenetic study on Sclerotinaceae showed that most of the sclerotial lineages including different species of Sclerotinia had evolved relatively recently from a common ancestor [@carbone1993ribosomal].
Few multilocus genotypes were unique to each sampling state of the USA but most of them were common among several states (Supplemental Table 1).
For instance, Michigan state from which we had the largest collection of isolates (N = 62) consisted of 43 MLGs.
Of these, 32 MLGs were present in at least one of the 5 states of NE, ND, WA, CO, MN; 4 MLGs were present in France isolates; and 11 were unique to MI sites (see Supplemental Table 1 for unique MLGs in different regions).
When selecting *S. sclerotiorum* isolates for screening dry bean germplasm for disease resistance, plant breeders may find above information useful.
A study on genetic diversity of *S. sclerotiorum* from multiple host crops in the North central region of the USA revealed there was little evidence for genetic isolation of this pathogen by host crop or by distance [@aldrich-wolfe2015genetic].
Though our isolates didn't come from different crops, we had isolates from many resistant and susceptible dry bean cultivars, which also didn't show genetic differentiation according to AMOVA.
It seems Mexican pathogen isolates are different from the rest and warrants further investigation using a larger population.
Some France genotypes were common in USA isolates but at the same time 43% (6/14) of the France genotypes were unique to that region indicating white mold resistant cultivars produced in the USA might not work in that region. <!-~ ZNK: I'm not entirely sold on this conclusion. ~->
Geographic distance between different states of the USA seem not a barrier for gene transfer of *S. sclerotiorum* isolates though unique genotypes did prevail at a low frequency in each state. <!-~ ZNK: I don't know that this adds anything. ~->
The genetic diversity of regions with >10 isolates had similar values (0.43 - 0.59) except for Mexican isolates.
*h* of MX was 0.26 and indicated less exposure of sampling sites to outside inocula. <!-~ ZNK: Does this mean that there's heavy inbreeding here? ~->
Test for linkage disequilibrium showed evidence of clonal reproduction for MX isolates (Ia = 2.61, P.Ia = 0.001, $\bar{r}_d$ = 0.37, P.$\bar{r}_d$  = 0.001).
This may have been another reason for low gene diversity.

DAPC analysis by region followed by AMOVA corroborated the uniqueness of Mexican genotypes by grouping them away from the rest of the isolates (Fig. 8).
Other populations based on geography were unable to group in separate clusters except for the Californian population.
We tested whether genotypes of *S. sclerotiorum* populations were influenced by different dry bean host cultivars and year of sampling by DAPC and AMOVA.
There was no host specificity but year of sampling did show a significant impact on genotypes.
Other studies also reported host identity is no barrier for gene flow of this pathogen [@aldrich-wolfe2015genetic; @clarkson2013population].
@atallah2004high revealed statistically significant impact of the collection date of *S. sclerotiorum* isolates from four potato fields in the Columbia Basin on molecular variance.
But there was no impact of the year or location of collection.
We also did not detect location as a significant predictor of molecular variance but year had a significant impact.
Our samples were collected on a single day in each year from each field and therefore did not warrant for an analysis of date of sample collection. <!-~ BSA: Is this true, coauthors? ~->

@@

*Sclerotinia sclerotiorum* isolates from dry bean represented high microsatellite diversity and lack of association MCGs and microsatellite MLGs.
Analysis of molecular variance of populations based on region and sampling year showed low level of differentiation with majority of variance detected within populations.
These features are indicative of sexual outcrossing and perhaps human mediated migration of *S. sclerotiorum* in dry bean.
However, we also detected clones or identical microsatellite MLGs within and among populations indicating asexual reproduction.
Further, linkage disequilibrium and index of association calculations suggested *S. sclerotiorum* populations are largely clonal.
Although, there was low impact of region and year on population differentiation, we did not detect host cultivar as a barrier to gene flow of the pathogen.
Our study did not find a widespread genotype among sampled regions from dry beans but some Australian and France MLGs were found in the USA isolates.
The analysis of populations from same field over different years for linkage disequilibrium did not provide evidence for sexual outcrossing.
However, analysis of molecular variance did show populations were significantly different between some years (eg.
MI 2003 and 2005 populations; WA 2008 compared to both 2003 and 2005).
These results indicate some level of heterothallic recombination is taking place in these fields in addition to natural mutations.
An understanding of sexual behavior and ability of form new genotypes by *S. sclerotiorum* in dry bean cropping systems is useful for formulating disease management strategies. -->
